---
title: "Minimal example: Wien, 2 scenarios, no LAI sweep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minimal example: Wien, 2 scenarios, no LAI sweep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

is_ghactions <- tolower(Sys.getenv("GITHUB_ACTIONS")) == "true" ||
  tolower(Sys.getenv("CI")) %in% c("true", "1", "yes")

path_base <- system.file("extdata/models/wien/base.h5", package = "kwb.raindrop")
data_available <- nzchar(path_base) && file.exists(path_base)
is_windows <- Sys.info()[["sysname"]] == "Windows"
```

This vignette is a small, self-contained reduction of the full Wien
workflow used for end-to-end checks (CI and review). It runs only
**two scenarios** (Daniel's reference design plus one mulde-area
variation) and intentionally does **not** sweep `LAI` so that the
default value baked into `base.h5` is left in place.

The companion full-grid vignette (`workflow_wien.Rmd`) varies LAI as
part of a 432-row parameter grid; see Michael's analysis (May 2026)
which showed that LAI between 3.9 and 8.5 changes the modelled ET
share by ~0.1 percentage points only.

### Input data

Precipitation (10-minute resolution) and evapotranspiration (daily
ET0) for 2011-2025 are provided by
[GeoSphere Austria](https://www.geosphere.at/) (Österreichischer
Wetterdienst, formerly ZAMG) and ship with the package under
`inst/extdata/models/wien/`. The HDF5 model template (`base.h5`)
is produced with the Tandler "Regenwasserbewirtschaftung" calculation
engine; the engine itself is downloaded from the
`KWB-R/kwb.raindrop.binaries` GitHub Release on demand via
`kwb.raindrop::download_engine()`.

### Static `base.h5` parameter overview

All values present in the shipped `base.h5` template. The two
scenarios below override only a small subset (`connected_area`,
`mulde_area`, `mulde_height`, `filter_hydraulicconductivity`,
filter/storage Schichtdicken, `Endversickerungsrate`, `LAI` is
*not* touched). Everything else listed here is the static default
that drives the simulation.

```{r show_static_params, eval = data_available}
library(kwb.raindrop)

h5 <- hdf5r::H5File$new(path_base, mode = "r")
all_vals <- kwb.raindrop::h5_read_values(h5)
h5$close_all()

format_value <- function(v) {
  if (is.data.frame(v)) {
    sprintf("[time series: %d rows, columns: %s]",
            nrow(v), paste(names(v), collapse = ", "))
  } else if (length(v) > 1L) {
    paste(format(v, trim = TRUE), collapse = ", ")
  } else {
    format(v, trim = TRUE)
  }
}

base_params <- tibble::tibble(
  parameter = names(all_vals),
  value = vapply(all_vals, format_value, character(1L))
) |>
  dplyr::arrange(parameter)

DT::datatable(
  base_params,
  filter = "top",
  options = list(pageLength = 25, autoWidth = TRUE),
  caption = sprintf("All %d parameters in base.h5 (Wien)", nrow(base_params))
)
```

### Two-scenario parameter grid

Daniel's reference design plus one variation (`mulde_area` doubled).

```{r preparation, eval = data_available}
path_list <- list(
  modelname = "Wien",
  root_path = file.path(tempdir(), "raindrop_wien_minimal"),
  dir_input  = "<root_path>/models/<modelname>/input",
  dir_output = "<root_path>/models/<modelname>/output",
  dir_target_output = "<dir_output>/<dir_target>",
  file_errors_hdf5 = "Fehlerprotokoll.h5",
  file_results_hdf5_element = "Mulde_Rigole.h5",
  file_results_hdf5_flaeche = "Dach.h5",
  file_results_hdf5_verschaltungen = "<dir_target>_Verschaltungen.h5",
  file_results_txt = "Mulde_Rigole_RAINDROP.txt",
  file_results_txt_multilayer = "Mulde_Rigole_RAINDROP_multi_layer.txt",
  file_target = "<dir_target>.h5",
  path_base = system.file("extdata/models/wien/base.h5",  package = "kwb.raindrop"),
  path_exe  = if (is_windows) kwb.raindrop::download_engine() else NA_character_,
  path_et   = system.file("extdata/models/wien/et.csv",   package = "kwb.raindrop"),
  path_rain = system.file("extdata/models/wien/rain.csv.gz", package = "kwb.raindrop"),
  path_errors_hdf5 = "<dir_target_output>/<file_errors_hdf5>",
  path_results_hdf5_element = "<dir_target_output>/<file_results_hdf5_element>",
  path_results_hdf5_flaeche = "<dir_target_output>/<file_results_hdf5_flaeche>",
  path_results_hdf5_verschaltungen = "<dir_target_output>/<file_results_hdf5_verschaltungen>",
  path_results_txt = "<dir_target_output>/<file_results_txt>",
  path_results_txt_multilayer = "<dir_target_output>/<file_results_txt_multilayer>",
  path_target_input = "<dir_input>/<file_target>"
)

param_grid <- tibble::tibble(
  scenario_name = c("s00001", "s00002"),
  connected_area               = 1000,
  mulde_area                   = c(75, 150),
  mulde_height                 = 200,
  filter_hydraulicconductivity = 36,
  filter_height                = 300,
  storage_height               = 500,
  bottom_hydraulicconductivity = 12,
  rain_factor                  = 1
)

DT::datatable(
  param_grid,
  options = list(pageLength = 5, autoWidth = TRUE),
  caption = "Variable parameters (everything not listed here stays at the base.h5 default)"
)

psi_s_mm <- function(kf_mmh) (3.237 * (kf_mmh / 25.4)^(-0.328)) * 25.4

paths <- kwb.utils::resolve(path_list)

timeseries_et <- readr::read_delim(paths$path_et, delim = ";", col_types = "cd") |>
  dplyr::mutate(
    date = lubridate::dmy(date),
    time = as.integer(difftime(date, min(date), units = "hours"))
  ) |>
  dplyr::select(-date) |>
  dplyr::filter(!is.na(value)) |>
  dplyr::relocate(time, .before = value)
timeseries_et$time[nrow(timeseries_et)] <- ceiling(timeseries_et$time[nrow(timeseries_et)])

timeseries_rain <- readr::read_csv(paths$path_rain, show_col_types = FALSE) |>
  dplyr::rename(datetime = time, value = rr) |>
  dplyr::mutate(time = as.double(difftime(datetime, min(datetime), units = "secs")) / 3600) |>
  dplyr::filter(!is.na(value)) |>
  dplyr::select(-datetime, -station) |>
  dplyr::relocate(time, .before = value)
timeseries_rain$time[nrow(timeseries_rain)] <- ceiling(timeseries_rain$time[nrow(timeseries_rain)])

# Align ET / rain end-times to the longer of the two
if (max(timeseries_rain$time) > max(timeseries_et$time)) {
  timeseries_et <- dplyr::bind_rows(
    timeseries_et,
    tibble::tibble(time = max(timeseries_rain$time),
                   value = timeseries_et$value[nrow(timeseries_et)])
  )
}
if (max(timeseries_et$time) > max(timeseries_rain$time)) {
  timeseries_rain <- dplyr::bind_rows(
    timeseries_rain,
    tibble::tibble(time = max(timeseries_et$time),
                   value = timeseries_rain$value[nrow(timeseries_rain)])
  )
}

# Convert rain from mm to mm/h (per 10-min interval)
period <- c(diff(timeseries_rain$time), mean(diff(timeseries_rain$time)))
timeseries_rain$value <- timeseries_rain$value / period
```

### Run the two scenarios

```{r run_model, eval = data_available && is_windows}
run_one <- function(i, timestep_hours, debug = FALSE, ...) {
  param_grid_tmp <- param_grid[i, ]
  paths <- kwb.utils::resolve(path_list, dir_target = param_grid_tmp$scenario_name)

  fs::dir_create(paths$dir_input, recurse = TRUE)
  fs::dir_create(paths$dir_output, recurse = TRUE)
  fs::dir_create(paths$dir_target_output, recurse = TRUE)

  fs::file_copy(path = paths$path_base,
                new_path = paths$path_target_input,
                overwrite = TRUE)

  h5 <- hdf5r::H5File$new(paths$path_target_input, mode = "a")
  new_path <- stringr::str_c(normalizePath(fs::path_abs(paths$dir_target_output)), "\\")
  vals <- kwb.raindrop::h5_read_values(h5)

  vals$`//Berechnungsparameter/Ergebnispfad` <- new_path
  vals$`//Berechnungsparameter/Zeitschritt_Infiltration` <- timestep_hours
  vals$`//Berechnungsparameter/Zeitschritt_ET` <- timestep_hours
  vals$`//Berechnungsparameter/Zeitschritt_Verschaltungen` <- timestep_hours
  vals$`//Berechnungsparameter/R-Plots` <- 0
  vals$`//Berechnungsparameter/Ausgabemodus` <- "Optimierung"
  vals$`//Berechnungsparameter/Evapotranspiration_aktiv` <- 1

  vals$`//Massnahmenelemente/Dach/Berechnungsparameter/Evapotranspiration_aktiv` <- 1
  vals$`//Massnahmenelemente/Dach/Allgemein/Flaeche` <- param_grid_tmp$connected_area

  vals$`//Massnahmenelemente/Mulde_Rigole/Berechnungsparameter/Evapotranspiration_aktiv` <- 1
  vals$`//Massnahmenelemente/Mulde_Rigole/Allgemein/Regen-Skalierungsfaktor` <- param_grid_tmp$rain_factor
  vals$`//Massnahmenelemente/Mulde_Rigole/Allgemein/Flaeche` <- param_grid_tmp$mulde_area
  vals$`//Massnahmenelemente/Mulde_Rigole/Eigenschaften_Oberflaeche/Ueberlaufhoehe` <- param_grid_tmp$mulde_height
  vals$`//Massnahmenelemente/Mulde_Rigole/Bodenschichtung/Startwerte_theta_ActualSoilMoisture` <- c(0.3, 0)
  vals$`//Massnahmenelemente/Mulde_Rigole/Bodenschichtung/Schichtdicken` <- c(
    param_grid_tmp$filter_height, param_grid_tmp$storage_height
  )
  vals$`//Massnahmenelemente/Mulde_Rigole/Allgemein/Endversickerungsrate` <-
    param_grid_tmp$bottom_hydraulicconductivity

  vals$`//Bodenarten/Bodenfilter/Ks_HydraulicConductivity` <- param_grid_tmp$filter_hydraulicconductivity
  vals$`//Bodenarten/Bodenfilter/Psi_Saugspannung_CapillarySuction` <-
    psi_s_mm(param_grid_tmp$filter_hydraulicconductivity)

  # NOTE: LAI is intentionally NOT overridden — base.h5 default is used.

  vals$`//Kurven/ET0` <- timeseries_et
  vals$`//Kurven/Regen` <- timeseries_rain
  vals$`//Kurven/Growth_1`$time[2] <- max(timeseries_rain$time)
  vals$`//Kurven/Shading_1`$time[2] <- max(timeseries_rain$time)

  kwb.raindrop::h5_write_values(h5, vals, resize = TRUE,
                               scalar_strategy = "error", verbose = FALSE)
  h5$close_all()

  kwb.raindrop::run_model(path_exe = paths$path_exe,
                          path_input = paths$path_target_input,
                          debug = debug)
  invisible(NULL)
}

kwb.raindrop::run_scenarios(
  indices = seq_len(nrow(param_grid)),
  run_one_scenario = run_one,
  timestep_hours = 0.1,
  debug = FALSE,
  parallel = FALSE,
  show_progress = FALSE
)
```

### Results

```{r analyse_results, eval = data_available && is_windows}
simulation_results <- kwb.raindrop::get_simulation_results_optim_parallel(
  paths = paths,
  path_list = path_list,
  simulation_names = param_grid$scenario_name,
  debug = FALSE
)

simulation_results_optimisation <- kwb.raindrop::add_overflow_events_and_waterbalance(
  simulation_results = simulation_results,
  event_separation_hours = 4
)

results <- param_grid |>
  dplyr::left_join(simulation_results_optimisation, by = c("scenario_name" = "s_name")) |>
  dplyr::relocate(scenario_name, .before = connected_area)

DT::datatable(
  results,
  options = list(pageLength = 5, autoWidth = TRUE),
  caption = "Two-scenario simulation results (water balance + overflow events)"
)
```
